---
title: "Complex Model Prior Sensitivity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(posterior))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(tidyr))

theme_set(theme_minimal())
color_scheme_set("mix-blue-red")
```

## Data preparation

```{r}
data_processed <- read.csv("../data/processed/oasis_processed.csv")

data <- data_processed %>%
  select(Age, Educ, nWBV, CDR, sex_male) %>%
  filter(
    !is.na(Age),
    !is.na(Educ),
    !is.na(nWBV),
    !is.na(CDR),
    !is.na(sex_male),
    Age > 0,
    nWBV > 0,
    sex_male %in% c(0, 1),
    CDR %in% c(0, 0.5, 1)
  ) %>%
  mutate(
    age_z = as.numeric(scale(Age)),
    brain_volume_z = as.numeric(scale(nWBV)),
    educ_z = as.numeric(scale(Educ)),
    brain_educ_z = brain_volume_z * educ_z,
    y_cat = case_when(
      CDR == 0   ~ 1L,
      CDR == 0.5 ~ 2L,
      CDR == 1   ~ 3L
    )
  )

stan_data_base <- list(
  N = nrow(data),
  y_cat = data$y_cat,
  sex = as.numeric(data$sex_male),
  age = data$age_z,
  brain_volume = data$brain_volume_z,
  educ = data$educ_z,
  brain_educ = data$brain_educ_z
)
```

## Stan model for prior sensitivity

The original complex Stan model fixes the priors. Here we write a temporary Stan file so the prior settings can be varied from R.

## Fit the model under several priors

```{r}
mod <- cmdstan_model("complex_model_sensitivity.stan")

prior_grid <- tibble(
  prior = c("beta N(0, 0.5)", "beta N(0, 1)", "beta N(0, 2)"),
  beta_sd = c(0.5, 1.0, 2.0),
  c1_sd = c(1.0, 1.0, 1.0),
  delta_rate = c(0.3, 0.3, 0.3)
)

fit_one_complex <- function(prior, beta_sd, c1_sd, delta_rate) {
  fit <- mod$sample(
    seed = 405,
    chains = 4,
    iter_sampling = 4000,
    refresh = 1000,
    show_messages = FALSE,
    data = c(
      stan_data_base,
      list(
        beta_sd = beta_sd,
        c1_sd = c1_sd,
        delta_rate = delta_rate
      )
    )
  )

  list(prior = prior, fit = fit)
}

fit_list <- pmap(prior_grid, fit_one_complex)
names(fit_list) <- prior_grid$prior
```

## Posterior summaries

```{r}
params <- c("beta1", "beta2", "beta3", "beta4", "beta34", "c1", "delta", "c2")

extract_summary <- function(fit_obj, prior_name) {
  draws <- as_draws_df(fit_obj$draws(params))

  tibble(
    parameter = params,
    mean = sapply(params, function(p) mean(draws[[p]])),
    median = sapply(params, function(p) median(draws[[p]])),
    sd = sapply(params, function(p) sd(draws[[p]])),
    q2.5 = sapply(params, function(p) quantile(draws[[p]], 0.025)),
    q97.5 = sapply(params, function(p) quantile(draws[[p]], 0.975)),
    prob_negative = sapply(params, function(p) mean(draws[[p]] < 0)),
    prior = prior_name
  )
}

summary_table <- map2_dfr(
  fit_list,
  names(fit_list),
  ~ extract_summary(.x$fit, .y)
)

summary_table %>%
  arrange(parameter, prior)
```

```{r}
summary_table %>%
  filter(parameter %in% c("beta3", "beta34", "c1", "delta", "c2")) %>%
  arrange(parameter, prior)
```

## Diagnostics

```{r}
diagnostic_table <- map2_dfr(
  fit_list,
  names(fit_list),
  function(x, nm) {
    x$fit$summary(params) %>%
      as_tibble() %>%
      mutate(prior = nm) %>%
      select(prior, variable, mean, sd, q5, q95, rhat, ess_bulk, ess_tail)
  }
)

diagnostic_table
```

## Posterior density plots

```{r}
posterior_beta3 <- map2_dfr(
  fit_list,
  names(fit_list),
  function(x, nm) {
    draws <- as_draws_df(x$fit$draws("beta3"))
    tibble(value = draws$beta3, prior = nm)
  }
)

ggplot(posterior_beta3, aes(x = value, color = prior)) +
  geom_density(linewidth = 1) +
  labs(
    title = "Posterior density of brain-volume effect across priors",
    x = expression(beta[3]),
    y = "Density",
    color = "Prior"
  )
```

```{r}
posterior_beta34 <- map2_dfr(
  fit_list,
  names(fit_list),
  function(x, nm) {
    draws <- as_draws_df(x$fit$draws("beta34"))
    tibble(value = draws$beta34, prior = nm)
  }
)

ggplot(posterior_beta34, aes(x = value, color = prior)) +
  geom_density(linewidth = 1) +
  labs(
    title = "Posterior density of interaction effect across priors",
    x = "beta34",
    y = "Density",
    color = "Prior"
  )
```

## Credible interval plots

```{r}
interval_plot_data <- summary_table %>%
  filter(parameter %in% c("beta3", "beta34", "c1", "delta", "c2"))

ggplot(
  interval_plot_data,
  aes(x = prior, y = median, ymin = q2.5, ymax = q97.5)
) +
  geom_pointrange() +
  facet_wrap(~ parameter, scales = "free_y") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Posterior medians and 95% credible intervals across priors",
    x = "Prior",
    y = "Posterior estimate"
  )
```

## Trace plots and rank histograms

```{r}
for (nm in names(fit_list)) {
  print(
    mcmc_trace(fit_list[[nm]]$fit$draws(c("beta3", "beta34", "c1", "delta", "c2"))) +
      ggtitle(paste("Trace plots:", nm)) +
      xlab("Iteration Number")
  )
}
```

```{r}
for (nm in names(fit_list)) {
  print(
    mcmc_rank_hist(fit_list[[nm]]$fit$draws(c("beta3", "beta34", "c1", "delta", "c2"))) +
      ggtitle(paste("Rank histograms:", nm))
  )
}
```

## Calibration plots by class and prior

```{r}
get_pred_means <- function(fit_obj) {
  list(
    pred_0 = colMeans(as_draws_matrix(fit_obj$draws("p1"))),
    pred_05 = colMeans(as_draws_matrix(fit_obj$draws("p2"))),
    pred_1 = colMeans(as_draws_matrix(fit_obj$draws("p3")))
  )
}

calibration_check <- function(pred_prob, observed_outcome, bins = 8) {
  df <- tibble(
    predicted = pred_prob,
    observed = observed_outcome
  ) %>%
    mutate(bin = ntile(predicted, bins))

  df %>%
    group_by(bin) %>%
    summarise(
      mean_predicted = mean(predicted),
      mean_observed = mean(observed),
      n = n(),
      .groups = "drop"
    )
}

# IMPORTANT: use the same filtered data used for modeling
observed_0 <- as.numeric(data$y_cat == 1L)
observed_05 <- as.numeric(data$y_cat == 2L)
observed_1 <- as.numeric(data$y_cat == 3L)

calibration_all <- map2_dfr(
  fit_list,
  names(fit_list),
  function(x, nm) {
    preds <- get_pred_means(x$fit)

    bind_rows(
      calibration_check(preds$pred_0, observed_0, bins = 8) %>% mutate(class = "0"),
      calibration_check(preds$pred_05, observed_05, bins = 8) %>% mutate(class = "0.5"),
      calibration_check(preds$pred_1, observed_1, bins = 8) %>% mutate(class = "1")
    ) %>%
      mutate(prior = nm)
  }
)

ggplot(calibration_all, aes(x = mean_predicted, y = mean_observed, group = 1)) +
  geom_point() +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  facet_grid(class ~ prior) +
  labs(
    title = "Calibration plots across priors and classes",
    x = "Mean predicted probability",
    y = "Observed proportion"
  )
```
