---
title: "complex_model"
output: html_document
---

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(posterior))
suppressPackageStartupMessages(library(loo))
```

We first load the processed data and select the necessary columns. 
```{r}
data_processed = read.csv("../../data/processed/oasis_processed.csv")

data_processed = data_processed |> select(Age, Educ, nWBV, CDR, sex_male)

head(data_processed)
```


```{r}
N = nrow(data_processed)
y = data_processed$CDR
age = data_processed$Age
sex = data_processed$sex_male
brain_volume =data_processed$nWBV
educ = data_processed$Educ
brain_educ = data_processed$Educ*data_processed$nWBV

## Standardising
age_z = as.numeric(scale(age))
brain_volume_z = as.numeric(scale(brain_volume))
educ_z = as.numeric(scale(educ))
brain_educ_z = brain_volume_z*educ_z


```

Now we compute the posterior using stan. 

```{r}
mod <- cmdstan_model("complex_model.stan")
```

```{r}
fit = mod$sample(
  seed =405,
  chains = 4,
  refresh = 1000,
  data = list( 
          y = y, 
          N = nrow(data_processed),
          sex = sex,
          age = age_z,
          brain_volume = brain_volume_z,
          educ = educ_z,
          brain_educ = brain_educ_z
        ), 
  iter_sampling = 4000,
  show_messages = FALSE
)

fit
fit_summary <- fit$summary()
write.csv(fit_summary, "../../results/complex_model/fit_summary.csv", row.names = FALSE)
```
Now, we assess the behaviour of our chains using trace plots. 

```{r}
summary_df <- fit$summary()
write.csv(summary_df, "fit_summary_complex_model.csv", row.names = FALSE)
```

```{r}
theme_set(theme_minimal())

p_beta1 <- mcmc_trace(fit$draws("beta1"))  + xlab("Iteration Number")
p_k0 <- mcmc_trace(fit$draws("k0"))     + xlab("Iteration Number")
p_delta <- mcmc_trace(fit$draws("delta"))  + xlab("Iteration Number")
p_beta34 <- mcmc_trace(fit$draws("beta34")) + xlab("Iteration Number")

p_beta1
p_k0
p_delta
p_beta34

ggsave("../../plots/plots_complex_model/beta1_trace.png", plot = p_beta1, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/k0_trace.png", plot = p_k0, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/delta_trace.png", plot = p_delta, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/beta34_trace.png", plot = p_beta34, width = 7, height = 4, dpi = 300)
```

We create a rank plot to inspect the rank distribution and mixing behavior.

```{r}

rank_hist_beta1 <- mcmc_rank_hist(fit$draws("beta1")) +
                      ylab("no of MCMC samples with these ranks") +
                      theme_minimal() + 
                      theme(axis.text.x = element_text(angle = 45, hjust = 1))

rank_hist_beta2 <- mcmc_rank_hist(fit$draws("beta2")) +
                      ylab("no of MCMC samples with these ranks") +
                      theme_minimal() + 
                      theme(axis.text.x = element_text(angle = 45, hjust = 1))

rank_hist_beta34 <- mcmc_rank_hist(fit$draws("beta34")) +
                      ylab("no of MCMC samples with these ranks") +
                      theme_minimal() + 
                      theme(axis.text.x = element_text(angle = 45, hjust = 1))

rank_hist_k0 <- mcmc_rank_hist(fit$draws("k0")) +
                  ylab("no of MCMC samples with these ranks") +
                  theme_minimal() + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1))

rank_hist_delta <- mcmc_rank_hist(fit$draws("delta")) +
                      ylab("no of MCMC samples with these ranks") +
                      theme_minimal() + 
                      theme(axis.text.x = element_text(angle = 45, hjust = 1))

rank_hist_beta1
rank_hist_beta2
rank_hist_beta34
rank_hist_k0
rank_hist_delta

ggsave("../../plots/plots_complex_model/beta1_rank.png", plot = rank_hist_beta1, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/beta2_rank.png", plot = rank_hist_beta2, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/beta34_rank.png", plot = rank_hist_beta34, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/k0_rank.png", plot = rank_hist_k0, width = 7, height = 4, dpi = 300)
ggsave("../../plots/plots_complex_model/k0_rank.png", plot = rank_hist_delta, width = 7, height = 4, dpi = 300)
  
```

```{r}

pred_0 = colMeans(as_draws_matrix(fit$draws("p0")))
pred_05= colMeans(as_draws_matrix(fit$draws("p05")))
pred_1 =  colMeans(as_draws_matrix(fit$draws("p1")))

```
We create function to get the data for the calibration plot. 
```{r}
calibration_check = function(pred_prob, observed_outcome, bins) {
  df = data.frame(predicted = pred_prob, 
                  observed = observed_outcome)
  
  df$bin = cut(df$predicted, breaks = bins)
  
  summary = df |> group_by(bin) |>
           summarise(mean_predicted = mean(predicted),
                     mean_observed = mean(observed), 
                     n = n())
        
  return(summary)
    
}

```

Now use the function to generate the calibtration plot data and then plot it
```{r}
observed_0 = as.numeric(data_processed$CDR == 0)
observed_1 = as.numeric(data_processed$CDR == 1)
observed_05= as.numeric(data_processed$CDR == 0.5)

c0 = calibration_check(pred_0, observed_0, bins = 8)
c1 = calibration_check(pred_1, observed_1, bins = 8)
c05 = calibration_check(pred_05, observed_05, bins = 8)

c0 
c1
c05
```
We create function so we can efficiently make the calibration plots. 

```{r}
calibration_plot = function(cal_df, title) {
  ggplot(cal_df, aes(x = mean_predicted, y = mean_observed)) + 
  geom_point() + 
  geom_line() +
  labs(x = "Mean of the predicted value", y = "Proportion of Observed") +
  geom_abline(intercept = 0, slope = 1, 
                linetype = "dashed", color = "red") +
  ggtitle(title) 
}

```

```{r}
p_c0  <- calibration_plot(c0, "Class 0")
p_c1  <- calibration_plot(c1, "Class 1")
p_c05 <- calibration_plot(c05, "Class 05")

p_c0
p_c1
p_c05

ggsave("../../plots/plots_complex_model/calibration_class0.png", plot = p_c0, width = 6, height = 5, dpi = 300)
ggsave("../../plots/plots_complex_model/calibration_class1.png", plot = p_c1, width = 6, height = 5, dpi = 300)
ggsave("../../plots/plots_complex_model/calibration_class05.png", plot = p_c05, width = 6, height = 5, dpi = 300)

```


We will check the predictive power of our model using leave-one-out prediction.

```{r}
log_lik <- fit$draws("log_lik", format = "matrix")

loo_result <- loo(log_lik)
elpd <- loo_result$estimates["elpd_loo", "Estimate"]

N <- nrow(data_processed)
avg_prob <- exp(elpd / N)

loo_output <- list(
  loo = loo_result,
  elpd = elpd,
  avg_prob = avg_prob
)

saveRDS(loo_output, "../../results/complex_model/loo_output.rds")

print(loo_result)
cat("Average probability assigned to true outcome:", round(avg_prob, 3), "\n")
```
The Pareto k estimates tell us that there are no outliers influencing the model. \newline
The model is assigning 45.3% probability to the correct outcome (correct class of dementia diagnosis) on average for the leave-one-out predictions.  This is worse than the naive model, telling us the extra complexity prevents the model from generalizing well.


```{r}
draws <- fit$draws() 

mcmc_hist(draws, pars = "beta1")
mcmc_hist(draws, pars = "beta2")
mcmc_hist(draws, pars = "beta3")
mcmc_hist(draws, pars = "beta34")


```

```{r}

samples <- as_draws_df(fit$draws("beta1"))$beta1
mean(samples >0)
samples34 <- as_draws_df(fit$draws("beta34"))$beta34
2*(1 -mean(samples34  >0))



```
```{r}
fit$draws() |>
  summarise_draws(
    mean,
    ~quantile2(.x, probs = c(0.05, 0.95))  # gives columns q5 and q95
  )

```
```{r}
beta3_draws  <- as.vector(fit$draws("beta3",  format = "matrix"))
beta43_draws <- as.vector(fit$draws("beta34", format = "matrix"))

effect_low_educ = beta3_draws +beta43_draws*(-1)
effect_mean_educ= beta3_draws +beta43_draws*(0)
effect_high_educ = beta3_draws + beta43_draws*(1)

summarise_effect <- function(draws) {
  c(mean    = mean(draws),
    lower95 = quantile(draws, 0.025),
    upper95 = quantile(draws, 0.975),
    prob_negative = mean(draws < 0)) 
}

summarise_effect(effect_low_educ)
summarise_effect(effect_mean_educ)
summarise_effect(effect_high_educ)


```

