---
title: "complex_model"
output: html_document
---

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(posterior))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(purrr))

theme_set(theme_minimal())
color_scheme_set("mix-blue-red")
```

We first load the processed data and select the necessary columns. 
```{r}
data_processed <- read.csv("../../data/processed/oasis_processed.csv")

data_processed <- data_processed |>
  select(Age, Educ, nWBV, CDR, sex_male) |>
  filter(
    !is.na(Age),
    !is.na(Educ),
    !is.na(nWBV),
    !is.na(CDR),
    !is.na(sex_male)
  )

N <- nrow(data_processed)
y <- data_processed$CDR
age <- data_processed$Age
sex <- data_processed$sex_male
brain_volume <- data_processed$nWBV
educ <- data_processed$Educ

# Standardise continuous predictors
age_z <- as.numeric(scale(age))
brain_volume_z <- as.numeric(scale(brain_volume))
educ_z <- as.numeric(scale(educ))
brain_educ_z <- brain_volume_z * educ_z

stan_data <- list(
  N = N,
  y = y,
  sex = sex,
  age = age_z,
  brain_volume = brain_volume_z,
  educ = educ_z,
  brain_educ = brain_educ_z
)
```
Now we compute the posterior using stan. 

```{r}
mod <- cmdstan_model("complex_model.stan")
fit_hmc = mod$sample(
  seed =405,
  chains = 4,
  refresh = 1000,
  data = stan_data,
  iter_sampling = 4000,
  show_messages = FALSE
)

fit_hmc$summary(c("beta1", "beta2", "beta3", "beta4", "beta34", "k0", "delta", "sigma"))
```
Fit Variational Inference with meanfield (alternatively try fullrank)
```{r}
fit_vi <- mod$variational(
  data = stan_data,
  seed = 405,
  algorithm = "meanfield",
  iter = 20000,
  grad_samples = 1,
  elbo_samples = 100,
  output_samples = 4000,
)

fit_vi$summary(c("beta1", "beta2", "beta3", "beta4", "beta34", "k0", "delta", "sigma"))
```

Extract posterior draws

```{r}
params <- c("beta1", "beta2", "beta3", "beta4", "beta34", "k0", "delta", "sigma")

draws_hmc <- as_draws_df(fit_hmc$draws(params))
draws_vi  <- as_draws_df(fit_vi$draws(params))
```

Comparison table

```{r}
make_summary <- function(draws_df, method_name, params) {
  data.frame(
    parameter = params,
    mean = sapply(params, function(p) mean(draws_df[[p]])),
    median = sapply(params, function(p) median(draws_df[[p]])),
    sd = sapply(params, function(p) sd(draws_df[[p]])),
    q2.5 = sapply(params, function(p) quantile(draws_df[[p]], 0.025)),
    q97.5 = sapply(params, function(p) quantile(draws_df[[p]], 0.975)),
    prob_negative = sapply(params, function(p) mean(draws_df[[p]] < 0)),
    method = method_name
  )
}

summary_comparison <- rbind(
  make_summary(draws_hmc, "HMC", params),
  make_summary(draws_vi, "VI meanfield", params)
)

summary_comparison
```

```{r}
summary_comparison %>%
  filter(parameter %in% c("beta3", "beta34")) %>%
  arrange(parameter, method)
```

Posterior deinsity comparison

Brain Volume effect

```{r}
plot_df_beta3 <- rbind(
  data.frame(value = draws_hmc$beta3, method = "HMC"),
  data.frame(value = draws_vi$beta3, method = "VI meanfield")
)

p_beta3 <- ggplot(plot_df_beta3, aes(x = value, colour = method)) +
  geom_density(linewidth = 1) +
  labs(
    title = "Posterior comparison for beta3",
    x = expression(beta[3]),
    y = "Density"
  )

p_beta3

ggsave("../../plots/complex_model/beta3_density_hmc_vi.png", plot = p_beta3, width = 6, height = 4, dpi = 300)
```
The posterior distributions obtained via HMC and mean-field variational inference (VI) are broadly similar in location but differ notably in dispersion. For the brain-volume coefficient $\beta_{3}$, the VI posterior is substantially more concentrated than the HMC posterior, indicating underestimation of uncertainty. This is a well-known limitation of mean-field VI, which assumes independence among parameters and therefore cannot capture posterior correlations.


Interval comparison plot

```{r}
plot_df_beta34 <- rbind(
  data.frame(value = draws_hmc$beta34, method = "HMC"),
  data.frame(value = draws_vi$beta34, method = "VI meanfield")
)

p_beta34 <- ggplot(plot_df_beta34, aes(x = value, colour = method)) +
  geom_density(linewidth = 1) +
  labs(
    title = "Posterior comparison for beta34",
    x = expression(beta[34]),
    y = "Density"
  )

p_beta34

ggsave("../../plots/complex_model/beta34_density_hmc_vi.png", plot = p_beta34, width = 6, height = 4, dpi = 300)
```
In contrast, the posterior distributions for the interaction term $\beta_{34}$ are much more closely aligned between the two methods. Both the location and spread are similar, suggesting that the marginal posterior for this parameter is closer to a Gaussian distribution and less affected by dependence structures.

The discrepancy observed for $\beta_3$ likely reflects underlying posterior dependence with other model parameters, such as the interaction term and intercept-related parameters. HMC captures the full joint posterior structure, including correlations, whereas mean-field VI approximates the posterior as fully factorized. As a result, VI compresses marginal variances when parameters are correlated.

The smaller discrepancy for $\beta_{34}$ suggests that this parameter is less entangled with others in the posterior geometry, making it easier for the factorized approximation to perform well.

```{r}
calibration_check <- function(pred_prob, observed_outcome, bins = 8) {
  df <- data.frame(
    predicted = pred_prob,
    observed = observed_outcome
  )

  df$bin <- dplyr::ntile(df$predicted, bins)

  df |>
    group_by(bin) |>
    summarise(
      mean_predicted = mean(predicted),
      mean_observed = mean(observed),
      n = n(),
      .groups = "drop"
    )
}

get_class_predictions <- function(fit_obj) {
  list(
    pred_0 = colMeans(as_draws_matrix(fit_obj$draws("p0"))),
    pred_05 = colMeans(as_draws_matrix(fit_obj$draws("p05"))),
    pred_1 = colMeans(as_draws_matrix(fit_obj$draws("p1")))
  )
}
```

```{r}
observed_0  <- as.numeric(data_processed$CDR == 0)
observed_05 <- as.numeric(data_processed$CDR == 0.5)
observed_1  <- as.numeric(data_processed$CDR == 1)
```


```{r}
preds_hmc <- get_class_predictions(fit_hmc)

c0_hmc  <- calibration_check(preds_hmc$pred_0, observed_0, bins = 8)
c05_hmc <- calibration_check(preds_hmc$pred_05, observed_05, bins = 8)
c1_hmc  <- calibration_check(preds_hmc$pred_1, observed_1, bins = 8)

c0_hmc$method <- "HMC"
c05_hmc$method <- "HMC"
c1_hmc$method <- "HMC"

c0_hmc$class <- "0"
c05_hmc$class <- "0.5"
c1_hmc$class <- "1"
```


```{r}
preds_vi <- get_class_predictions(fit_vi)

c0_vi  <- calibration_check(preds_vi$pred_0, observed_0, bins = 8)
c05_vi <- calibration_check(preds_vi$pred_05, observed_05, bins = 8)
c1_vi  <- calibration_check(preds_vi$pred_1, observed_1, bins = 8)

c0_vi$method <- "VI meanfield"
c05_vi$method <- "VI meanfield"
c1_vi$method <- "VI meanfield"

c0_vi$class <- "0"
c05_vi$class <- "0.5"
c1_vi$class <- "1"
```


```{r}
calibration_all <- rbind(
  c0_hmc, c05_hmc, c1_hmc,
  c0_vi, c05_vi, c1_vi
)

ggplot(calibration_all, aes(x = mean_predicted, y = mean_observed)) +
  geom_point() +
  geom_line() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  facet_grid(class ~ method) +
  labs(
    title = "Calibration plots: HMC vs VI",
    x = "Mean predicted probability",
    y = "Observed proportion"
  )
```

```{r}
fit_hmc$summary(params)
```
```{r}
mcmc_trace(fit_hmc$draws(c("beta1", "beta3", "beta34", "k0", "delta", "sigma"))) +
  xlab("Iteration Number")
```
```{r}
mcmc_rank_hist(fit_hmc$draws(c("beta1", "beta3", "beta34", "k0", "delta", "sigma")))
```

```{r}
fit_vi
```

