---
title: "Naive Model"
output: pdf_document
---

```{r setup, include=FALSE}
# imports 
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r}
library(dplyr)
library(cmdstanr)
library(ggplot2)
library(bayesplot)
library(loo)
```

First, lets import the data and prep it to be fed into the stan model.
```{r}
data_processed = read.csv("../../data/processed/oasis_processed.csv")

# selecting relevant columns and enforcing feature constraints
data <- data_processed %>%
  select(sex_male, Age, nWBV, Educ, CDR_binary) %>%
  filter(Educ >= 1,
         Educ <= 5,
         Age > 0,
         nWBV > 0,
         (sex_male == 1 | sex_male == 0),
         (CDR_binary == 1 | CDR_binary == 0)
         )

# creating the data object to be passed into stan
stan_data <- list(
  N = nrow(data),
  sex_male = data$sex_male,
  age = data$Age,
  brain_vol = data$nWBV,
  educ = data$Educ,
  CDR_binary = data$CDR_binary
)
```

Next, lets retrieve the naive model from the stan file and fit it.
```{r}
# get stan model
naive_model <- cmdstan_model("naive_model.stan")

fit <- naive_model$sample(
  seed = 405,
  data = stan_data,
  chains = 4,
  iter_sampling = 2500,
  show_messages = FALSE
)
```

```{r}
fit_summary <- fit$summary()
write.csv(fit_summary, "fit_summary.csv", row.names = FALSE)
```
`rhat` $\approx 1$ for all parameters, indicating that the chains have all converged to the same distribution.  The bulk and tail ESS are high, indicating efficient sampling.  

#### Calibration Plot

Let's perform a goodness of fit check on the naive model.
```{r}
# extract probabilities from fitted model object
draws_df <- fit$draws(format = "df")
p_cols <- grep("^p_pred\\[", names(draws_df), value = TRUE)
p_matrix <- as.matrix(draws_df[, p_cols])
p_mean <- colMeans(p_matrix)
```

```{r}
calibration <- data.frame(
  y_true = data$CDR_binary,
  p = p_mean
) %>%
  mutate(bins = ntile(p, 15)) %>%
  group_by(bins) %>%
  summarise(
    mean_pred = mean(p),
    mean_true = mean(y_true)
  )

ggplot(calibration, aes(x = mean_pred, y = mean_true)) +
  geom_point() +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal() +
  xlab("Mean Predicted Probability") + 
  ylab("Mean True Value (Diagnosis)") +
  ggtitle("Calibration Plot")

ggsave("plots/naive_model/calibration.png")
```
As we can see, the plot is not very calibrated.  Therefore, we should try standardizing the numeric predictors to see if we get a better result.

```{r}
# selecting relevant columns and enforcing feature constraints
data <- data_processed %>%
  select(sex_male, Age, nWBV, Educ, CDR_binary) %>%
  filter(Educ >= 1,
         Educ <= 5,
         Age > 0,
         nWBV > 0,
         (sex_male == 1 | sex_male == 0),
         (CDR_binary == 1 | CDR_binary == 0)
         ) %>%
  mutate(Age_standardized = as.numeric(scale(Age)),
         nWBV_standardized = as.numeric(scale(nWBV)),
         Educ_standardized = as.numeric(scale(Educ)))

# creating the data object to be passed into stan
stan_data <- list(
  N = nrow(data),
  sex_male = data$sex_male,
  age = data$Age_standardized,
  brain_vol = data$nWBV_standardized,
  educ = data$Educ_standardized,
  CDR_binary = data$CDR_binary
)

fit <- naive_model$sample(
  seed = 405,
  data = stan_data,
  chains = 4,
  iter_sampling = 2500,
  show_messages = FALSE
)

# extract probabilities from fitted model object
draws_df <- fit$draws(format = "df")
p_cols <- grep("^p_pred\\[", names(draws_df), value = TRUE)
p_matrix <- as.matrix(draws_df[, p_cols])
p_mean <- colMeans(p_matrix)

calibration <- data.frame(
  y_true = data$CDR_binary,
  p = p_mean
) %>%
  mutate(bins = ntile(p, 15)) %>%
  group_by(bins) %>%
  summarise(
    mean_pred = mean(p),
    mean_true = mean(y_true)
  )

ggplot(calibration, aes(x = mean_pred, y = mean_true)) +
  geom_point() +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal() +
  xlab("Mean Predicted Probability") + 
  ylab("Mean True Value (Diagnosis)") +
  ggtitle("Calibration Plot")
```
That did not help much.  Therefore, we should try other fixes in the complex model implementation if it's necessary.

#### Rank Plots, Trace Plots, Mixing Plots

First, let's create a rank plot to inspect the rank distribution and mixing behavior.
```{r}
mcmc_rank_hist(fit$draws("beta_0")) +
  ylab("no of MCMC samples with these ranks") +
  ggtitle("MCMC Ranks Histogram for Prior Beta 0") +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(0, 10000, by = 2500)
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.ticks.x = element_line()
  )
```
```{r}
mcmc_rank_hist(fit$draws("beta_3")) +
  ylab("no of MCMC samples with these ranks") +
  ggtitle("MCMC Ranks Histogram for Prior Beta 3") +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(0, 10000, by = 2500)
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.ticks.x = element_line()
  )
```

```{r}
mcmc_rank_hist(fit$draws("beta_4")) +
  ylab("no of MCMC samples with these ranks") +
  ggtitle("MCMC Ranks Histogram for Prior Beta 4") +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(0, 10000, by = 2500)
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.ticks.x = element_line()
  )
```

```{r}
mcmc_rank_hist(fit$draws("beta_34")) +
  ylab("no of MCMC samples with these ranks") +
  ggtitle("MCMC Ranks Histogram for Interaction Term Prior") +
  theme_minimal() +
  scale_x_continuous(
    breaks = seq(0, 10000, by = 2500)
  ) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    axis.ticks.x = element_line()
  )
```

All of the histograms look roughly flat for all 4 chains for the parameters displayed.  This is an indication of good mixing.

Next, we will create some trace plots.
```{r}
mcmc_trace(fit$draws("beta_0")) + theme_minimal() +
  xlab("Iteration Number") +
  ylab("Value of Beta 0") +
  ggtitle("Trace Plot for Beta 0 Prior")
```
```{r}
mcmc_trace(fit$draws("beta_3")) + theme_minimal() +
  xlab("Iteration Number") +
  ylab("Value of Beta 3") +
  ggtitle("Trace Plot for Beta 3 Prior")
```

```{r}
mcmc_trace(fit$draws("beta_4")) + theme_minimal() +
  xlab("Iteration Number") +
  ylab("Value of Beta 4") +
  ggtitle("Trace Plot for Beta 4 Prior")
```

```{r}
mcmc_trace(fit$draws("beta_34")) + theme_minimal() +
  xlab("Iteration Number") +
  ylab("Value of Beta 34") +
  ggtitle("Trace Plot for Interaction Prior")
```
All 4 chains have plenty of overlap, and don't appear to be stuck at any point.  This demonstrates good mixing for all of the priors shown. 

Next, we will make autocorrelation plots.
```{r}
mcmc_acf(fit$draws("beta_0")) +
  ggtitle("Autocorrelation Plot for Prior Beta 0")
```

```{r}
mcmc_acf(fit$draws("beta_3")) +
  ggtitle("Autocorrelation Plot for Prior Beta 3")
```

```{r}
mcmc_acf(fit$draws("beta_4")) +
  ggtitle("Autocorrelation Plot for Prior Beta 4")
```

```{r}
mcmc_acf(fit$draws("beta_34")) +
  ggtitle("Autocorrelation Plot for Interaction Prior")
```
All priors show fast decay to 0 for all chains, demonstrating efficient sampling.

### Leave-One-Out Prediction
Now we will try leave-one-out prediction on our fitted model to see how it performs.
```{r}
log_lik <- fit$draws("log_lik", format = "matrix")

loo_result <- loo(log_lik)

elpd <- loo_result$estimates["elpd_loo", "Estimate"]

N <- nrow(data)
avg_prob <- exp(elpd / N)

print(loo_result)
cat("Average probability assigned to true outcome:", round(avg_prob, 3), "\n")
```
The Pareto k estimates tell us that there are no outliers influencing the model. \newline
The model is assigning 57.3% probability to the correct outcome (correct class of dementia diagnosis) on average for the leave-one-out predictions.  