---
title: "complex_model"
output: html_document
---

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(posterior))

```

We first load the processed data and select the necessary columns. 
```{r}
data_processed = read.csv("../data/processed/oasis_processed.csv")

data_processed = data_processed |> select(Age, Educ, nWBV,CDR, sex_male )
head(data_processed)


```
```{r}
N = nrow(data_processed)
y = data_processed$CDR
age = data_processed$Age
sex = data_processed$sex_male
brain_volume =data_processed$nWBV
educ = data_processed$Educ
brain_educ = data_processed$Educ*data_processed$nWBV

## Standardising
age_z = as.numeric(scale(age))
brain_volume_z = as.numeric(scale(brain_volume))
educ_z = as.numeric(scale(educ))
brain_educ_z = brain_volume_z*educ_z
```

Now we compute the posterior using stan. 

```{r}

mod <- cmdstan_model("complex_model.stan")
fit = mod$sample(
  seed =405,
  chains = 4,
  refresh = 1000,
  data = list( 
          y = y, 
          N = nrow(data_processed),
          sex = sex,
          age = age_z,
          brain_volume = brain_volume_z,
          educ = educ_z,
          brain_educ = brain_educ_z
        ), 
  iter_sampling = 4000
)


```

```{r}
fit
```
Now, we assess the behaviour of our chains using trace plots. 

```{r}

theme_set(theme_minimal())

mcmc_trace(fit$draws("beta1"))  + xlab("Iteration Number")
mcmc_trace(fit$draws("k0"))     + xlab("Iteration Number")
mcmc_trace(fit$draws("delta"))  + xlab("Iteration Number")
mcmc_trace(fit$draws("beta34")) + xlab("Iteration Number")
```

We create a rank plot to inspect the rank distribution and mixing behavior.

```{r}

mcmc_rank_hist(fit$draws("beta1")) +
  ylab("no of MCMC samples with these ranks") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mcmc_rank_hist(fit$draws("beta2")) +
  ylab("no of MCMC samples with these ranks") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mcmc_rank_hist(fit$draws("beta34")) +
  ylab("no of MCMC samples with these ranks") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mcmc_rank_hist(fit$draws("k0")) +
  ylab("no of MCMC samples with these ranks") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

mcmc_rank_hist(fit$draws("delta")) +
  ylab("no of MCMC samples with these ranks") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

  
```

```{r}

pred_0 = colMeans(as_draws_matrix(fit$draws("p0")))
pred_05= colMeans(as_draws_matrix(fit$draws("p05")))
pred_1 =  colMeans(as_draws_matrix(fit$draws("p1")))



```
We create function to get the data for the calibration plot. 
```{r}
calibration_check = function(pred_prob, observed_outcome, bins) {
  df = data.frame(predicted = pred_prob, 
                  observed = observed_outcome)
  
  df$bin = cut(df$predicted, breaks = bins)
  
  summary = df |> group_by(bin) |>
           summarise(mean_predicted = mean(predicted),
                     mean_observed = mean(observed), 
                     n = n())
        
  return(summary)
    
}

```

Now use the function to generate the calibtration plot data and then plot it
```{r}
observed_0 = as.numeric(data_processed$CDR == 0)
observed_1 = as.numeric(data_processed$CDR == 1)
observed_05= as.numeric(data_processed$CDR == 0.5)

c0 = calibration_check(pred_0, observed_0, bins = 8)
c1 = calibration_check(pred_1, observed_1, bins = 8)
c05 = calibration_check(pred_05, observed_05, bins = 8)

c0 
c1
c05
```
We create function so we can efficiently make the calibration plots. 

```{r}
calibration_plot = function(cal_df, title) {
  ggplot(cal_df, aes(x = mean_predicted, y = mean_observed)) + 
  geom_point() + 
  labs(x = "Mean of the predicted value", y = "Proportion of Observed") +
  geom_abline(intercept = 0, slope = 1, 
                linetype = "dashed", color = "red") +
  ggtitle(title) 
}

```

```{r}
calibration_plot(c0, "Class 0")

calibration_plot(c1, "Class 1")

calibration_plot(c05, "Class 05")

```



```