---
title: "simulation"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2)
library(cmdstanr)
library(posterior)
library(tidyverse)
```

First read the data and standardise predictors. Use only a random subset of the data for correctness of code checks. 
```{r}
data_processed = read.csv("../data/processed/oasis_processed.csv")

data_processed = data_processed |> select(Age, Educ, nWBV,CDR, sex_male)
data_processed 

predictors = data_processed |> select(-CDR) |> mutate(across(Age: nWBV, scale)) |> slice_sample(n = 80)

simulation_df <- data.frame(
  variable    = character(),
  lower       = numeric(),
  upper       = numeric(),
  true_value  = numeric(),
  recovered   = logical(), 
  iteration = numeric()
)
simulation_df
```


We generate parameters and data according to our model many times. Then after fitting our model, what proportions of the times the 90%credible interval for the parameter is able to capture the true value of the parameter. 
```{r}
Niter = 30

for (i in 1:Niter) {

sd_beta = 2

beta1 = rnorm(1,mean =0,sd = sd_beta)
beta2 = rnorm(1, mean = 0, sd_beta)
beta3 = rnorm(1, mean = 0, sd_beta)
beta4 = rnorm(1, mean = 0, sd_beta)
beta34 = rnorm(1, mean = 0, sd_beta)
delta = rexp(1, 0.3)
sigma = rexp(1, 0.3)
k0 = rnorm(1, 0,1)

params = c(beta1, beta2, beta3, beta4,beta34, delta, sigma, k0)

get_disease = function(age, educ, nWBV, sex_male) {
  mu = beta1*sex_male +beta2*age + beta3*nWBV +beta4*educ +beta34*nWBV*educ
  Z = rnorm(1, mu, sd = sigma)
  if (Z <= k0) {
    return(0)
  } else if (Z <= k0 + delta) {
    return(0.5)
  } else {
    return(1)
  }
}

full_data <- predictors |>
  rowwise() |>
  mutate(y = get_disease(Age, Educ, nWBV, sex_male)) |>
  ungroup()

full_data

y = as.numeric(full_data$y)
age = as.numeric(full_data$Age)
sex = as.numeric(full_data$sex_male)
brain_volume =as.numeric(full_data$nWBV)
educ = as.numeric(full_data$Educ)
brain_educ = as.numeric(educ*brain_volume)

mod <- cmdstan_model("complex_model.stan")
fit = mod$sample(
  seed =405,
  chains = 1,
  refresh = 1000,
  data = list( 
          y = y, 
          N = nrow(full_data),
          sex = sex,
          age = age,
          brain_volume = brain_volume,
          educ = educ,
          brain_educ = brain_educ
        ), 
  iter_sampling = 3000,
  show_messages = FALSE
)



true_values = c(beta1 = beta1,
                beta2 = beta2,
                beta3 = beta3,
                beta4 =beta4,
                beta34 = beta34,
                delta = delta,
                sigma = sigma,
                k0 =k0)

draws = fit$draws(variables = names(true_values))

summary_df = summarise_draws(
  draws,
  mean,
  sd,
  rhat,
  ess_bulk,
  ~quantile(., probs = c(0.025, 0.975))
)

summary_df = summary_df |> rename(lower= `2.5%`, upper = `97.5%`) |>
  mutate(true_value = true_values[variable], 
         recovered= true_value <= upper & true_value >= lower, iteration = rep(i, 8))

simulation_df = bind_rows(simulation_df, summary_df)

}

```
```{r}
simulation_results = simulation_df |> select(iteration, recovered, variable) |> mutate(recovered = as.numeric(recovered)) |> group_by(variable) |> summarise(prob_recovered = mean(recovered))
simulation_results
```
We can see that for each parameter of the model proportion of the times each it is in the 90% credible interval is roughly 90%, so this confirms that our code can work as intended. 